Shallow EDSLs and Object-Oriented Programming

### 2017-02-20 paper/paper.tex

\documentclass[10pt,preprint,numbers,nocopyrightspace]{sigplanconf}

@@

\usepackage{amsmath}
\usepackage{color}
\usepackage{listings}
\usepackage{balance}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{pstricks,pst-node,pst-tree}
\usepackage{theorem}
\usepackage{bbm}
\usepackage{pgf}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{mdwlist}
\usepackage{morefloats}
\usepackage[T1]{fontenc}
\usepackage[scaled=0.85]{beramono}
\usepackage{mathptmx}
\usepackage[normalem]{ulem}
\usepackage{booktabs}
\usepackage{tablefootnote}
\usepackage{enumitem}
\usepackage{stmaryrd}
\usepackage{syntax}

\setlist[itemize]{noitemsep,nolistsep}
\setlist[enumerate]{noitemsep,nolistsep}
\setlength{\floatsep}{6.0pt plus 2.0pt minus 2.0pt}
\setlength{\textfloatsep}{10.0pt plus 2.0pt minus 4.0pt}
\setlength{\intextsep}{6.0pt plus 2.0pt minus 2.0pt}

\newcommand{\name}{\textbf{@Family}\xspace}
\newcommand{\dsl}{\emph{scans}\xspace}
\newcommand{\interp}{\textsc{Interpreter}\xspace}

@@

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
\copyrightyear{20yy}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\copyrightdoi{nnnnnnn.nnnnnnn}

@@

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{Shallow EDSLs and Object-Oriented Programming}   % 'preprint' option specified.

\title{Shallow EDSLs and Object-Oriented Programming}

\authorinfo{Weixin Zhang \and Bruno C.d.S.Oliveira}
           {The University of Hong Kong, Pokfulam, Hong Kong, China}
           {\{wxzhang2,bruno\}@cs.hku.hk}
%\authorinfo{Name}
%           {Affiliation2/3}
%           {Email2/3}

@@

%===============================================================================
\input{sections/Introduction.tex}
\input{sections/ShallowOO.tex}
\input{sections/Interpretations.tex}
\input{sections/Transformations.tex}
%\input{sections/RelatedWorkShort.tex}
\input{sections/Conclusion.tex}
%\input{sections/Modularity.tex}
%\input{sections/Overview.tex}
%\input{sections/Tool.tex}
%\input{sections/CaseStudy.tex}
%\input{sections/RelatedWork.tex}
%\input{sections/Discussion.tex}
%\acks

>>>>

\documentclass{jfp1}

@@

\usepackage{color}
\usepackage{listings}
\usepackage{url}
\usepackage{xspace}
\usepackage{syntax}
\usepackage{comment}

\newcommand{\dsl}{\emph{scans}\xspace}
\newcommand{\interp}{\textsc{Interpreter}\xspace}

@@

@@

\title{Shallow EDSLs and Object-Oriented Programming}
\subtitle{(Functional Pearl)}

\author[W. Zhang and B. C. d. S. Oliveira]
        {WEIXIN ZHANG and BRUNO C. d. S. OLIVEIRA\\
         The University of Hong Kong, Hong Kong\\
         \email{bruno@cs.hku.hk}}
%\authorinfo{Name}
%           {Affiliation2/3}
%           {Email2/3}

@@

%===============================================================================
\input{sections/Introduction.tex}
\input{sections/ShallowOO.tex}
\input{sections/Interpretations.tex}
%\input{sections/Transformations.tex}
%\input{sections/RelatedWorkShort.tex}
\input{sections/Conclusion.tex}
%\input{sections/Modularity.tex}
%\input{sections/Overview.tex}
%\input{sections/Tool.tex}
%\input{sections/CaseStudy.tex}
%\input{sections/RelatedWork.tex}
%\input{sections/Discussion.tex}
%\acks

### 2017-02-20 paper/sections/Interpretations.tex

\subsection{Discussion}
Gibbons and Wu claim that in shallow
embeddings new language constructs are easy to add, but new
interpretations are hard. As our OOP approach shows, in OOP both
language constructs and new interpretations are easy to add in shallow
embeddings. In other words, the circuit DSL presented so far does not
suffer from the Expression Problem. The key point is that procedural
abstraction combined with OOP features (subtyping, inheritance and
type-refinement) adds expressiveness over traditional procedural
abstraction. Gibbons and Wu do discuss a number of advanced techniques that 
can solve some of the modularity problems. For example, using type
classes, \emph{finally
  tagless}~\cite{carette2009finally} can deal with the example in
Section~\ref{subsec:multiple}. However
tuples are still needed 
to deal with dependent interpretations. In contrast the approach
proposed here is just straightforward OOP, and dependent
interpretations are not a problem.
\begin{comment}
and \emph{data types a la
  carte}~\cite{swierstra2008data} (DTC).
Finally tagless approach uses a type class to abstract over all interpretations
of the language. Concrete interpretations are given through creating a data type and
making it an instance of that type. However, it forces dependent interpretations to be defined along with what they depend on.
DTC represents language constructs separately and composes them together using
extensible sums. However, not like OO languages which come with subtyping, one
has to manually implement the subtyping machinery for variants.
\end{comment}
Gibbons and Wu also show some different variants of interpretations,
such as context-sensitive interpretations. 
These interpretations are unproblematic as well.
Implementation details can be found online.

>>>>

\subsection{Context-Sensitive Interpretations}
Interpretations may rely on some mutable contexts.
Consider an interpretation that simplifies the representation of a circuit.
A circuit can be divided horizontally into layers.
Each layer can be represented as a sequence of pairs $(i,j)$, denoting the connection from wire $i$ to wire $j$.
For instance, circuit
$$above\ (beside\ (fan\ 2)\ (fan\ 2))\ (stretch\ [2,2]\ (fan\ 2))$$
can be represented as:
$$[[(0,1), (2,3)], [(1,3)]]$$
It has two layers: the first layer has connections from
the first wire to the second, and from the second to the third; the second layer has
only one connection from the first wire to the third one.

\paragraph{Context-sensitive Interpretations in Haskell}
The following Haskell code models the interpretation described above:
\lstinputlisting[linerange=2-18]{./code/ContextSensitive.hs}%APPLY:TLAYOUT_HS
\lstinline{tlayout} is firstly a dependent interpretation, relying on the
previously defined \lstinline{width} interpretation.
More importantly, it is a context-sensitive interpretation.
The domain of \lstinline{tlayout} is not a value but a function \lstinline{(Int -> Int) -> Layout}.
The argument \lstinline{f} is also a function that shifts the index of a wire.
\lstinline{f} varies in recursive calls for accumulating transformations
on wires and hence is context-sensitive.

\paragraph{Context-sensitive Interpretations in Scala}
Context-sensitive interpretations in our OO approach are unproblematic as well.
Mutable contexts can be captured as arguments of methods:
\begin{lstlisting}
type Layout = List[List[Tuple2[Int,Int]]]

def lzw[A](xs: List[A], ys: List[A])(f: (A, A) => A): List[A] = (xs, ys) match {
  case (Nil,_) => ys
  case (_,Nil) => xs
  case (x::xs,y::ys) => f(x,y)::lzw(xs,ys)(f)
}
def partialSum(ns: List[Int]): List[Int] = ns.scanLeft(0)(_ + _).tail

trait Circuit extends width.Circuit {
  def tlayout(f: Int => Int): Layout
}
trait Id extends Circuit with width.Id {
  def tlayout(f: Int => Int) = List()
}
trait Fan extends Circuit with width.Fan {
  def tlayout(f: Int => Int) = List(for (i <- List.range(1,n)) yield (f(0),f(i)))
}
trait Above extends Circuit with width.Above {
  val c1, c2: Circuit
  def tlayout(f: Int => Int) = c1.tlayout(f) ++ c2.tlayout(f)
}
trait Beside extends Circuit with width.Beside {
  val c1, c2: Circuit
  def tlayout(f: Int => Int) =
    lzw (c1.tlayout(f), c2.tlayout(f.andThen(c1.width + _))) (_ ++ _)
}
trait Stretch extends Circuit with width.Stretch {
  val c: Circuit
  def tlayout(f: Int => Int) = c.tlayout(f.andThen(partialSum(ns)(_) - 1))
}
\end{lstlisting}
\subsection{Discussion}
Gibbons and Wu claim that in shallow
embeddings new language constructs are easy to add, but new
interpretations are hard. As our OOP approach shows, in OOP both
language constructs and new interpretations are easy to add in shallow
embeddings. In other words, the circuit DSL presented so far does not
suffer from the Expression Problem. The key point is that procedural
abstraction combined with OOP features (subtyping, inheritance and
type-refinement) adds expressiveness over traditional procedural
abstraction. Gibbons and Wu do discuss a number of advanced techniques that 
can solve some of the modularity problems. For example, using type
classes, \emph{finally
  tagless}~\cite{carette2009finally} can deal with the example in
Section~\ref{subsec:multiple}. However
tuples are still needed 
to deal with dependent interpretations. In contrast the approach
proposed here is just straightforward OOP, and dependent
interpretations are not a problem.
\begin{comment}
and \emph{data types a la
  carte}~\cite{swierstra2008data} (DTC).
Finally tagless approach uses a type class to abstract over all interpretations
of the language. Concrete interpretations are given through creating a data type and
making it an instance of that type. However, it forces dependent interpretations to be defined along with what they depend on.
DTC represents language constructs separately and composes them together using
extensible sums. However, not like OO languages which come with subtyping, one
has to manually implement the subtyping machinery for variants.
\end{comment}
Gibbons and Wu also show some different variants of interpretations,
such as context-sensitive interpretations. 
These interpretations are unproblematic as well.
Implementation details can be found online.

### 2017-02-20 paper/sections/Introduction.tex

>>>>

In this paper, however, we prefer to interpret shallow embeddings
differently. We argue that a straightforward OO approach is really a
shallow embedding. The reason is that shallow embeddings are 
implemented using \emph{procedural abstraction} and, as Cook~\cite{cook09abstraction} argued,
procedural abstraction is the essence of Object-Oriented
Programming. Thus the implementation of a shallow
DSL in OOP languages should simply correspond to a standard
object-oriented program.

The main goal of this paper is to argue that OOP languages have
advantages for the implementation of shallow embeddings. This should
not come as a surprise, as OOP languages have evolved over more than 50
years to improve the use of procedural abstraction.

In order to understand how OOP can help with shallow embeddings, lets
first review some of the limitations commonly
found in the literature:

\begin{enumerate}

\item {\bf Single Interpretation} An often stated limitation of
  shallow embeddings is that they only support a single
  interpretation.

\item {\bf No Transformations} Another commonly stated limitation 
of shallow embeddings is that they do not support transformations,
preventing optimizations and other useful transformations.

\end{enumerate}

\noindent Both limitations are often used as motivators to switch to deep embeddings.

We show that OOP abstractions, including \emph{(multiple)
  inheritance}, \emph{subtyping} and \emph{type-refinement}, are
helpful to address those problems. For the first problem, we can
employ a recently proposed design pattern~\cite{eptrivially16}, which provides a simple
solution to the \emph{Expression Problem}~\cite{expPb} in OOP languages. Thus
using just standard OOP mechanims enables \emph{multiple modular
  interpretations} to co-exist and be combined in shallow embeddings.
For the second problem, we show that transformations can be encoded 
with recursive objects. Interestingly enough, for transformations it
is possible to port back the OO solution to
Haskell.

We make our arguments by taking a recent paper by Gibbons and Wu~\cite{gibbons2014folding},
where procedural abstraction is used in Haskell to model a simple
EDSL, and we recode that EDSL in Scala. Although from the
\emph{syntactical} point of view there are minor inconveniences in
the Scala version, from the \emph{semantic} and \emph{modularity} point of view the
Scala version has clear advantages.

### 2017-02-20 paper/sections/Overview.tex

The new constructs are: \emph{stretch ns c} which inserts additional wires into the circuit \emph{c} by
summing up \emph{ns} and $above\ c_1\ c_2$ which combines two circuits of the same width vertically.
Their types are given below:
\lstinputlisting[language=haskell,linerange=39-40]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES

@@

\lstinputlisting[language=haskell,linerange=39-40]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES
\emph{stretch ns c} inserts additional wires into the circuit \emph{c} by
summing up \emph{ns} and $above c_1 c_2$ combines two circuits of the same width vertically.
With these new constructs, more complex circuits can be constructed.
For example, Figure~\ref{sec:oo} shows the circuit constructed by \emph{stretch [2,2,2] (fan 3) `beside` fan 1}.

To accomplish this task in Haskell, we can simply define two more cases,
\texttt{stretch} and \texttt{above}, for the semantic function:
\lstinputlisting[language=haskell,linerange=44-45]{./code/shallowCircuit.hs}%APPLY:SYNTAX_HS
Similarly, defining two new interfaces
\texttt{Stretch} and \texttt{Above} that both extend \texttt{Circuit} and
implement the \texttt{width} method is all we need to do in Java:

>>>>

The new constructs are: \emph{stretch ns c} which inserts additional wires into the circuit \emph{c} by
summing up \emph{ns} and $above\ c_1\ c_2$ which combines two circuits of the same width vertically.
Their types are given below:
\lstinputlisting[language=haskell,linerange=29-30]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES

@@

\lstinputlisting[language=haskell,linerange=29-30]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES
\emph{stretch ns c} inserts additional wires into the circuit \emph{c} by
summing up \emph{ns} and $above c_1 c_2$ combines two circuits of the same width vertically.
With these new constructs, more complex circuits can be constructed.
For example, Figure~\ref{sec:oo} shows the circuit constructed by \emph{stretch [2,2,2] (fan 3) `beside` fan 1}.

To accomplish this task in Haskell, we can simply define two more cases,
\texttt{stretch} and \texttt{above}, for the semantic function:
\lstinputlisting[language=haskell,linerange=34-35]{./code/shallowCircuit.hs}%APPLY:SYNTAX_HS
Similarly, defining two new interfaces
\texttt{Stretch} and \texttt{Above} that both extend \texttt{Circuit} and
implement the \texttt{width} method is all we need to do in Java:

### 2017-02-20 paper/sections/RelatedWorkShort.tex

\paragraph{Deep and Shallow Embeddings}
Shallow embedding yields flexible and concise EDSLs while deep embedding makes
it easy to define optimizations.
There are a lot of work~\cite{svenningsson2012combining,
  Jovanovic:2014:YCD:2658761.2658771, scherr2014implicit} trying to blend these two
approaches to enjoy benefits from both.
They typically encode the surface language with a shallow embedding and
then generate or translate to a deep embedded version for allowing optimizations.
%Hofer and Ostermann~\cite{hofer2010modular} propose to provide both embedding through implementing internal and external visitor at the same time so that clients can choose for a particular interpretation;
The OO approach we present retains the simplicity of shallow embedding while
makes it possible to implement optimizations.

\paragraph{Modularity of EDSLs}
Languages that solves the EP are capable of being host languages for
implementing modular EDSLs. But many of these solutions require sophisticated
type parameterization, advanced features or heavy encoding which impose
complexity to the DSL implementers or even clients, preventing practical use.
For example, solutions based on the \textsc{Visitor}
pattern~\cite{oliveira09modular,hofer2010modular} %more references
Relatively simple solution like Object Algebras~\cite{bruno12oa} may not support
optimizations well.
Our approach, however, is simple and supports transformations.

>>>>

\paragraph{Deep and Shallow Embeddings}

### 2017-02-20 paper/sections/Transformations.tex

\subsection{Desugaring}
Desugaring is a transformation which eliminates 
language constructs without sacrificing the expressiveness of the language.
The \emph{id} construct is an example of syntactic sugar:
\vspace{-8pt}
$$id\ n = \overbrace{\ beside\ (fan\ 1)}^{\text{repeat }n-1\text{ times}}\ (fan\ 1)$$
The formula states that \emph{id n} can be rewritten as \emph{n} of \emph{fan 1}
combined with \emph{beside}.

\paragraph{Desugaring in Scala}
A corresponding Scala implementation is given below:
\lstinputlisting[linerange=4-17]{../src/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
The method \lstinline{desugar} returns a
\lstinline{Circuit} reflecting the nature of transformations.
Implementing each case is straightforward:
\emph{fan n} returns itself; $beside\ c_1\ c_2$ returns a new
$beside$ with $c_1$ and $c_2$ desugared; \emph{id n} mimics the
formula.

@@

%\weixin{define a show function in the and show the result after
%desugaring?}
\paragraph{Desugaring in Haskell}
Interestingly we can port the implementation back to Haskell:
\lstinputlisting[linerange=6-15]{./code/Desugar.hs}%APPLY:DESUGAR
\noindent This code follows well-known encodings of objects into 
\emph{functional} calculi with recursive records~\cite{bruce97comparing}. Thanks 
to Haskell's laziness, the port of recursive definitions such as
\lstinline{beside} is straightforward.

>>>>

\subsection{Desugaring}
Desugaring is a transformation which eliminates 
language constructs without sacrificing the expressiveness of the language.
The \emph{id} construct is an example of syntactic sugar:
\vspace{-5pt}
$$id\ n = \overbrace{\ beside\ (fan\ 1)}^{\text{repeat }n-1\text{
    times}}\ (fan\ 1)$$
\noindent The formula states that \emph{id n} can be rewritten as \emph{n} of \emph{fan 1}
combined with \emph{beside}.

\vspace{-3pt}
\paragraph{Desugaring in Scala}
A corresponding Scala implementation is given below:
\lstinputlisting[linerange=4-17]{../src/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
The method \lstinline{desugar} returns a
\lstinline{Circuit} reflecting the nature of transformations.
Implementing each case is straightforward:
\emph{fan n} returns itself; $beside\ c_1\ c_2$ returns a new
$beside$ with $c_1$ and $c_2$ desugared; \emph{id n} mimics the
formula.

@@

%\weixin{define a show function in the and show the result after
%desugaring?}
\vspace{-3pt}\paragraph{Desugaring in Haskell}
Interestingly we can port the implementation back to Haskell:
\lstinputlisting[linerange=6-15]{./code/Desugar.hs}%APPLY:DESUGAR
\noindent This code follows well-known encodings of objects into 
\emph{functional} calculi with recursive records~\cite{bruce97comparing}. Thanks 
to Haskell's laziness, the port of recursive definitions such as
\lstinline{beside} is straightforward.
